---
title: "PeakGuideR workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PeakGuideR workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview
PeakGuideR has been developed to help interpret complex MALDI-MSI peak datasets, 
where many detected m/z features may correspond to isotopes, adducst, 
matrix-related signals or alternative ion forms of the same underlying molecule.

IThe main idea of the workflos is to combine several proves of evidence that are 
often checked separately (see what the peaks are telling us):

- isotope morphology;
- carbon isotope-ratio support;
- elemental isotope-pattern support;
- adduct-family grouping;
- neutral-mass inference;
- database candidate matching and adduct database validation.

The output should therefore be read as annotation support rather than as final
compound identification. In practice, the package is intended to help prioritise
which peaks and candidate annotations are worth inspecting or validating further.

# Input data

This vignette shows the typical workflow starting from an already processed peak
matrix. The code chunks are not evaluated during package checks, so `pkm` and
`msi_peaks` should be replaced by the user's own objects.

The workflow can be run from 2 types of input:

1. a PeakGuideR peak matrix object or
2. a Cardinal MSI object that can be converted using `cardinal_to_peakmatrix()`.

For the peak-matrix route, the object should contain at least:

- `mass`: m/z values of detected features;
- `intensity`: pixel-by-feature intensity matrix.

For example:

Load the example peak matrix:

```{r load-example-pkm}
library(PeakGuideR)

data("example_pkm", package = "PeakGuideR")

pkm <- example_pkm

str(pkm)
```

# Running the workflow


For a standard analysis, the workflow can be run with only the peak matrix, the
ionisation mode and, optionally, the MALDI matrix:.

```{r run-workflow}
library(PeakGuideR)

res <- run_peakguider_workflow(
  pkm = example_pkm,
  ion_mode = "pos", #or "neg" for negative ionization
  matrix = "HCCA",#or NULL if it is not HCCA (until the moment, only HCCA matrix has been included in the library)
  quiet=TRUE)
```

Use `ion_mode = "neg"` for negative-ion mode data. The `matrix` argument is used
only for matrix-specific standard-adduct support. Currently, PeakGuideR includes
standard-adduct support for HCCA+DEA (ionic matrix); use `matrix = NULL` when this evidence layer
should not be used (recommended if you are using another matrix).

If the data are stored as a Cardinal MSI object, the same workflow can be called using `cardinal_object`:

```{r run-cardinal}
res <- run_peakguider_workflow(
  pkm = msi_peaks,
  ion_mode = "pos",
  matrix = "HCCA"
)
```

Internally, Cardinal objects are converted to peak-matrix format
using `cardinal_to_peakmatrix()`.

The returned object is a list with class `peakguider_workflow`:

```{r output-overview}
class(res)
names(res)
```

The workflow stores both intermediate evidence tables and final summary tables:

- `morph_results`: candidate isotope relationships supported by spatial morphology;
- `cir_results`: C13 isotope-ratio evidence for candidate isotope pairs;
- `eips_results`: elemental isotope-pattern evidence for non-carbon elements;
- `adduct_edges`: pairwise adduct relationships between features;
- `adduct_families`: groups of features compatible with the same inferred neutral mass;
- `relation_table`: a unified table collecting the feature-feature relationships;
- `feature_summary`: one row per detected m/z feature;
- `neutral_mass_candidates`: inferred neutral masses with database candidate matches.

# Feature-level summary

`feature_summary` contains one row per detected m/z feature.

This table is usually the first object to inspect, because it gives a feature-by-feature
view of the evidence found by the workflow. It reports whether each feature has
support for:

- being a monoisotopic peak with C13 support;
- being a C13 M+1 or M+2 isotope peak;
- having elemental isotope-pattern support;
- belonging to an adduct family;
- contributing to an inferred neutral mass.

```{r feature-summary}
res$feature_summary |>
  dplyr::select(
    idx,
    mz,
    is_c13_m0,
    c13_m1_idx,
    c13_score,
    has_c13_m2_support,
    has_eips,
    eips_elements,
    has_adduct_family,
    adduct_roles,
    main_adduct_role,
    neutral_mass_consensus
  ) |>
  head()
```

Useful columns include:

- `idx`: feature index.
- `mz`: feature m/z value.
- `is_c13_m0`: whether the feature is supported as a monoisotopic peak.
- `c13_m1_idx`: index of the putative C13 M+1 partner.
- `c13_score`: carbon isotope-ratio support score.
- `has_c13_m2_support`: whether C13 M+2 support was detected.
- `has_eips`: whether elemental isotope-pattern support was detected.
- `eips_elements`: elements supporting the EIPS evidence.
- `has_adduct_family`: whether the feature belongs to an inferred adduct family.
- `adduct_roles`: inferred adduct role or roles for the feature.
- `main_adduct_role`: main adduct role assigned from the strongest adduct family.
- `neutral_mass_consensus`: inferred neutral mass associated with the main adduct family.

A feature can have more than one type of evidence. For example, the same m/z
feature may be interpreted as a monoisotopic peak with C13 support and also as a member of an adduct family.

C13 evidence should be interpreted as relationship-based evidence. In simple
cases, PeakGuideR reports a candidate monoisotopic feature (`M+0`) and its
supported C13 isotopic features (`M+1` or `M+2`). In chained C13 patterns, a
feature may be involved in more than one isotope relationship, for example as
part of an `M+0 -> M+1 -> M+2` series. These cases are flagged with
`is_chained_c13` and should be inspected in `relation_table` before assigning a
final interpretation.
# Relation table

`relation_table` is useful when we want to see the individual relationships that led to the final feature-level summary.

It collects relationships detected by different modules into a common table.

```{r relation-table}
res$relation_table |>
  dplyr::select(
    from_idx,
    to_idx,
    from_mz,
    to_mz,
    relation_type,
    evidence_type,
    evidence_score,
    is_valid,
    group_id,
    from_role,
    to_role
  ) |>
  head()
```

Useful columns include:

- `from_idx`, `to_idx`: feature indices connected by the relationship.
- `from_mz`, `to_mz`: m/z values of the connected features.
- `relation_type`: type of relationship, such as C13, EIPS or adduct relation.
- `evidence_type`: evidence layer supporting the relationship.
- `evidence_score`: score associated with the relationship.
- `is_valid`: whether the relationship passed the selected threshold.
- `group_id`: group identifier, for example an adduct family ID.
- `from_role`, `to_role`: interpreted roles of the connected features.

This table is useful for tracing why a given feature was assigned a particular
role in `feature_summary`.

# Adduct families

`adduct_families` groups pairwise adduct relationships into families of features that are 
compatible with the same inferred neutral mass.

```{r adduct-families-names}
names(res$adduct_families)
```

The object contains three components:

- `family_summary`: one row per inferred adduct family;

- `family_members`: features assigned to each family and their inferred adduct roles;

- `family_edges`: pairwise adduct relationships supporting each family.

```{r adduct-family-summary}
res$adduct_families$family_summary |>
  head()

res$adduct_families$family_members |>
  head()

res$adduct_families$family_edges |>
  head()
```




# Peak relationship network

The relation table can also be visualised as a feature network with `plot_peak_network()`.

In this plot, nodes represent m/z features and edges represent relationships
supported by the workflow, such as isotope, EIPS or adduct relationships.

```{r peak-network}
plot_peak_network(
  relation_table = res$relation_table
)
```

By default, the function plots valid relationships with an evidence score of at
least 0.3. This threshold can be increased to focus on stronger relationships:

```{r peak-network-strict}
plot_peak_network(
  relation_table = res$relation_table,
  min_score = 0.5
)
```

The optional `idx_focus` argument can be used to inspect the neighbourhood of a
specific feature:

```{r peak-network-focus}
plot_peak_network(
  relation_table = res$relation_table,
  idx_focus = 25
)
```

This visualisation is exploratory. Connected features should be interpreted as
putative relationships, not as definitive annotations.

```{r peak-network-score}
plot_peak_network(
  relation_table = res$relation_table,
  min_score = 0.3,
  only_valid = TRUE
)
```

The optional `idx_focus` argument can be used to visualise the neighbourhood of
a specific feature:

```{r peak-network-focus-score}
plot_peak_network(
  relation_table = res$relation_table,
  min_score = 0.6,
  idx_focus = 25,
  only_valid = TRUE
)
```

Useful arguments include:

- `relation_table`: relation table returned by `run_peakguider_workflow()`.
- `min_score`: minimum evidence score required to draw an edge.
- `idx_focus`: optional feature index used to focus the network around one peak.
- `only_valid`: whether to plot only relationships that passed the validity criteria.
- `show_edge_labels`: whether to show relationship labels on the network edges.

The network should be interpreted as an exploratory visualisation. Connected
features are not necessarily definitive annotations; they represent putative
relationships supported by the evidence layers used by PeakGuideR.


# Neutral-mass candidates

`neutral_mass_candidates` is the main candidate annotation table.

It contains one row per inferred neutral mass and candidate compound match.

```{r neutral-mass-candidates}
res$neutral_mass_candidates |>
  dplyr::select(
    neutral_mass_id,
    neutral_mass_consensus,
    inferred_adducts,
    candidate_source,
    candidate_db_id,
    candidate_name,
    candidate_formula,
    candidate_neutral_mass,
    candidate_ppm_error,
    has_standard_compound_match,
    has_standard_adduct_match
  ) |>
  dplyr::filter(!is.na(candidate_name)) |>
  head()
```

Useful columns include:

- `neutral_mass_id`: internal neutral-mass group identifier.
- `neutral_mass_consensus`: inferred neutral mass.
- `inferred_adducts`: adducts supporting the inferred neutral mass.
- `candidate_source`: source database of the candidate match.
- `candidate_db_id`: database identifier.
- `candidate_name`: candidate compound name.
- `candidate_formula`: candidate molecular formula.
- `candidate_neutral_mass`: neutral monoisotopic mass from the database.
- `candidate_ppm_error`: mass error between the inferred neutral mass and the candidate.
- `has_standard_compound_match`: whether the candidate compound is present in the standard-adduct library.
- `has_standard_adduct_match`: whether the inferred adduct is supported for that compound in the standard-adduct library.

The database match is mass-based and should be interpreted as a putative
candidate, not as definitive identification.

# Summarising candidate matches

A quick summary of database and standard-adduct support can be obtained with:

```{r candidate-summary}
res$neutral_mass_candidates |>
  dplyr::summarise(
    n_rows = dplyr::n(),
    n_neutral_masses = dplyr::n_distinct(neutral_mass_id),
    n_rows_with_candidate = sum(!is.na(candidate_name)),
    n_neutral_masses_with_candidate = dplyr::n_distinct(
      neutral_mass_id[!is.na(candidate_name)]
    ),
    n_with_standard_compound_match = sum(has_standard_compound_match %in% TRUE),
    n_with_standard_adduct_match = sum(has_standard_adduct_match %in% TRUE)
  )
```

# Using full databases from Zenodo

PeakGuideR includes small example databases for testing and demonstration.

For real analyses, users should download the full non-commercial databases from
Zenodo: <https://doi.org/10.5281/zenodo.20705395>



After downloading the files, they can be loaded manually replacing the paths below with the real local paths

```{r load-full-dbs, eval=FALSE}
compound_db <- readRDS("path/to/compound_mass_database")
standards_db <- readRDS("path/to/adducts_database")


```

or using the helper function (also replace the file paths below with the local paths where the downloaded database files were saved.:

```{r load-full-dbs-helper, eval=FALSE}
dbs <- load_peakguider_databases(
  compound_db_path = "path/to/compound_mass_database.rds",
  standards_db_path = "path/to/standards_adduct_database.rds"
)
```

Then pass them to the workflow:

```{r workflow-full-dbs}
res <- run_peakguider_workflow(
  pkm = pkm,
  ion_mode = "pos",
  matrix = "HCCA",
  compound_db = dbs$compound_db,
  standards_db = dbs$standards_db
)
```


# Interpretation

PeakGuideR combines multiple evidence layers to support annotation.

However:

- isotope or adduct evidence does not prove compound identity,
- database matches are mass-based candidates,
- standard-adduct support is matrix- and preparation-specific,
- definitive identification may require MS/MS, standards or orthogonal validation.

The recommended interpretation is to use PeakGuideR outputs to prioritise
candidate annotations for downstream validation.
